---
classoption: table, dvipsname, compress
bibliography:  ref_GAMBAS.bib
biblio-style: "apalike"
output: 
  beamer_presentation:
    # keep_tex: true
    # toc: true
    slide_level: 3
    includes:
      in_header: header.tex
      before_body: doc_prefix.tex
      after_body: doc_suffix.tex
colorlinks: yes
fontsize: 10pt
header-includes: 
  - \usepackage{hyperref}
    \hypersetup{
      hidelinks,
      colorlinks=true,
      urlcolor=blue,
      linkcolor=black,
      breaklinks=true,
    }
link-citations: yes
linkcolor: Black
urlcolor: Maroon
---


```{r options, include=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(echo=TRUE, cache=TRUE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```

# jSDM package for fitting JSDMs

## Joint Species Distribution Models (JSDMs)
### Joint Species Distribution Models (JSDMs)

**Species Distribution Model (SDM), for one single species.**

- $\red{y_i} \sim \mathcal{B}ernoulli( \blue{\theta_i} )$
- $i$: site
- $g(\theta_i) = \red{X_i} \blue{\beta}$
- $\red{X}$: environmental variables
- $\blue{\beta}$: species effects

\begin{center}
\includegraphics[width=0.6\textwidth]{figs/SDM.jpg}
\end{center}

**JSDM $=$ SDM for community of species.**

### Joint Species Distribution Models (JSDMs)

**Model definition**

- $\red{y_{ij}}  \sim \mathcal{B}inomial(\red{n_i} , \blue{\theta_{ij}})$, for presence-absence data,
\vspace{0.5mm}
- $\red{y_{ij}} \sim\mathcal{P}oisson(\blue{\theta_{ij}})$, for abundance data, such that: 
\vspace{-0.7mm}
   $$g(\blue{\theta_{ij}})= \blue{\alpha_i} + \red{X_i}\blue{\beta_j} + \blue{W_i\lambda_j}$$
- $i$: site, $j$: species
- $g$: a link function (probit, logit or log).
- $\red{n_i}$: number of visits to site $i$.
- $\blue{\theta_{ij}}$: occurrence probability or mean abundance of species $j$ at site $i$.
- $\blue{\alpha_i}$: site effect (mean site suitability)
- $\red{X_i}$: environmental variables including the intercept.  
- $\blue{\beta_j}$: species effect  for species $j$,
- $\blue{W_i}$: latent variables (or "unmeasured predictors") for site $i$,
- $\blue{\lambda_j}$: factor loadings for species $j$
- $\blue{\Sigma_{jg}}=\blue{\lambda_j^t\lambda_g}$: coefficient for species $j$ and $g$ of the variance-covariance matrix  (species co-occurrences) used to compute the **full species residual correlation matrix**:
  $\blue{R_{jg}}:=\frac{\blue{\Sigma_{jg}}}{\sqrt{\blue{\Sigma _{jj}\Sigma _{gg}}}}.$

### Joint Species Distribution Models

JSDMs provide a convenient statistical framework to test **trait-environment** interactions.
\vspace{2mm}
\bcols
\bcol{0.7\textwidth}

$\beta_j$ can be expressed as a function of functional traits

- if $n$ species traits are considered:
    $\blue{\beta_j} \sim \mathcal{N}_{p+1}(\blue{\mu_{\beta_j}},V_{\beta})$,
    * with $\blue{\mu_{\beta_{jk}}}  = \sum_{r=0}^{n}  \blue{T_{jr}}.\blue{\gamma_{rk}}$, for $k \in \llbracket 0,p \rrbracket$, considering $p$ covariates. 
    * where $\blue{\gamma}=\bordermatrix{ & Int & X_1 & \ldots & X_p \cr
      Int & \blue{\gamma_{0,0}} & \blue{\gamma_{0,1}} & \ldots & \blue{\gamma_{0,p+1}} \cr
      T_1 & \blue{\gamma_{1,0}} & \blue{\gamma_{1,1}} & \ldots & \blue{\gamma_{1,p+1}} \cr
      \vdots & \dots & \ldots & \ldots & \ldots \cr
      T_n & \blue{\gamma_{n,0}} & \blue{\gamma_{n,1}} & \ldots & \blue{\gamma_{n+1,p+1}}}$
      
\vspace{2mm}   

- else:  $\blue{\beta_j} \sim \mathcal{N}_{p+1}(\mu_{\beta},V_{\beta})$, for $j \in  \llbracket 1,J \rrbracket$.

\ecol 

\bcol{0.3\textwidth}

\centering 

\includegraphics[width=0.9\textwidth]{figs/four-corner-model}
\ecol

\ecols

JSDMs can help narrow the gap between **correlative** and **mechanistic** species distribution models.

## `jSDM` R package

### `jSDM` R package

\includegraphics[width=0.89\textwidth]{figs/jSDM-website-2023-03.png}

- <https://ecology.ghislainv.fr/jSDM>
- Made with Rcpp* packages

### Available functions

The package includes the following functions to fit various species distribution models :

```{r jSDM-functions, echo=FALSE}
library(kableExtra)
names <- c("jSDM\\_binomial\\_logit", "jSDM\\_binomial\\_probit",
           "jSDM\\_binomial\\_probit\\_sp\\_constrained",
           "jSDM\\_binomial\\_probit\\_long\\_format",
           "jSDM\\_poisson\\_log", "jSDM\\_gaussian")
url <- paste0("https://ecology.ghislainv.fr/jSDM/reference/", names, ".html")
href <- paste0("\\href{", url, "}{", names, "}")
results <- data.frame(names=href,
                      type=c(rep("presence-absence", 4), "abundance", "continuous"),
                      format=c(rep("wide", 3), "long", rep("wide", 2)))
colnames(results) <- c("function", "data type", "data format")
knitr::kable(results, row.names=F, digits=1, format = 'latex', escape=FALSE,
             booktabs=TRUE, align = 'l') %>%
  kableExtra::kable_styling(latex_options=c("HOLD_position","striped"), full_width=FALSE) 
```

# Examples of jSDM usage

## Mapping tree communities in Madagascar 

### Mapping tree communities in Madagascar 

#### Data-sets considered

\vspace{-0.3cm}
\bcols
\bcol{0.4\textwidth}

- National forest inventories (1994-1996).

\centering
\includegraphics[width=0.65\textwidth]{figs/inv_sites.png}
\ecol
\bcol{0.6\textwidth}
\vspace{-0.2cm}
\hspace{-5mm} 

- Climatic variables (1960-1990):
    * Average annual temperature ($^ \circ C$),
    * Average annual precipitation (mm),
    * Seasonality of temperature,
    * Seasonality of precipitation,
    * Annual climatic water deficit (mm)
    
\centering
\vspace{0.1cm}
\includegraphics[width=1\textwidth]{figs/current-environmental-variables-1.png}

\ecol
\ecols

### Mapping tree communities in Madagascar 

#### Fitting JSDM from Madagascar data-sets

We performed, a quadratic binomial regression, using `jSDM`, such that: 
   $$\textrm{probit}(\blue{\theta_{ij}}) = \red{X_i}\blue{\beta_j} + \blue{W_i\lambda_j} + \blue{\alpha_i}.$$
   \vspace{4mm}
  **Data-sets dimensions, computation time and accuracy of results:**
  \includegraphics[width=0.96\textwidth]{figs/JSDM_Mada_poster.pdf}
  
### Mapping tree communities in Madagascar 

#### Residual correlation matrix estimated 

\bcols
\bcol{0.3\textwidth}
\vspace{3mm}
**Full species residual correlation matrix** computed from the estimated **variance-covariance matrix**: $$\blue{\Sigma_{jg}} = \blue{\lambda_j^T \lambda_{g}}$$,  
for $j,g \in  \llbracket 1,J \rrbracket$, as follows:
\vspace{3mm}
  $$\blue{R_{jg}}:=\frac{\blue{\Sigma_{jg}}}{\sqrt{\blue{\Sigma _{jj}\Sigma _{gg}}}}$$
\ecol
\bcol{0.7\textwidth}
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/res-corr-poster.eps}   
  
$R$ estimated between the 25 more abundant species. 
\end{center}
\ecol
\ecols

### Mapping tree communities in Madagascar 

#### Spatial interpolation of sites' parameters

\bcols
\bcol{0.6\linewidth} 

- **RST**: Regularized Spline with Tension using [`GRASS GIS`](https://grass.osgeo.org/) software via `rgrass7` R package [@Mitasova1993].
\vspace{5mm}
- **Occurrence probabilities** at inventory sites  $\Rightarrow$ **species distribution maps** at Madagascar scale

\ecol
\hspace{-3mm}
\bcol{0.4\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{figs/dist_sp1-poster.eps}
Estimated distribution of species *Ocotea laevis* and observed occurrences.

\ecol
\ecols

### Estimated species richness in Madagascar 
\vspace{-1mm}
- Species richness observed at site $i$ defined by $R_i:=\sum\limits_{j=1}^Jy_{ij}$.
\vspace{-1mm}
- Species richness estimated at site $i$ [@Scherrer2020]: $\widehat{R}_i=\sum\limits_{j=1}^J \hat{\theta}_{ij}$.
\vspace{-7mm}
\bcols
\hspace{-12mm}
\bcol{0.5\textwidth} 
\begin{center}
\includegraphics[width=1.1\textwidth]{figs/sp_rich_obs_fitted.pdf}  
\end{center}  
\vspace{-2mm}
Species richness estimated and observed at inventory sites.
\ecol
\hspace{-7mm}
\bcol{0.5\textwidth} 
\vspace{-8mm}
\begin{center}
\includegraphics[width=1.29\textwidth]{figs/sp-rich-deforest.pdf} 
\end{center}  
\vspace{-1mm}
Species richness estimated at Madagascar scale restricted to forest cover in 2000.
\ecol
\ecols

### Tree species community map in Madagascar
\vspace{-1mm}
**Method** [@Allnutt2008]:
\vspace{-1mm}

- Normalized PCA performed on species' occurrence probabilities,
- Coordinates of the PCA's first three axes scaled to [0,255],
- \textcolor{red}{R}\textcolor{ForestGreen}{G}\textcolor{blue}{B} coloration of the pixels given their scaled PCA's coordinates. 
\vspace{-5mm}
\bcols
\bcol{0.31\textwidth}
\begin{center}
\includegraphics[width=1.55\textwidth]{figs/beta-div.pdf}  
Species turnover estimated.
\end{center}
\ecol
\bcol{0.31\textwidth} 
\begin{center}
\includegraphics[width=1.5\textwidth]{figs/beta-div-deforest.pdf}  
\end{center}
\vspace{-2mm}
Estimated species turnover restricted to forest cover in 2000.
\ecol
\hspace{3mm}
\bcol{0.31\textwidth} 
\begin{center}
\includegraphics[width=0.67\textwidth]{figs/mada_forest_type.jpg} 
\end{center}
Forest types in Madagascar
\ecol
\ecols


### Tree species community map in Madagascar

- pixels with $\neq$ colors $\Rightarrow$ species present are not the same. 
- pixels with $\simeq$ colors $\Rightarrow$ communities of similar species.

\bcols
\bcol{0.31\textwidth}
\begin{center}
\includegraphics[width=1.55\textwidth]{figs/beta-div.pdf}  
Species turnover estimated.
\end{center}
\ecol
\bcol{0.31\textwidth} 
\begin{center}
\includegraphics[width=1.5\textwidth]{figs/beta-div-deforest.pdf}  
\end{center}
\vspace{-2mm}
Estimated species turnover restricted to forest cover in 2000.
\ecol
\hspace{3mm}
\bcol{0.31\textwidth} 
\begin{center}
\includegraphics[width=0.67\textwidth]{figs/mada_forest_type.jpg} 
\end{center}
Forest types in Madagascar
\ecol
\ecols

## Classification of tree communities in New Caledonia 

### Classification of tree communities in New Caledonia 
\vspace{-7mm}
#### Fitting JSDM in New Caledonia
\hspace{-5mm}
\bcols
\vspace{-7mm}
\bcol{0.5\textwidth}

**Explanatory variables** of the model and the orders considered: 

\vspace{2mm}
\includegraphics[height=0.37\textheight]{figs/X_NC.png}
\vspace{6mm}
\hspace{-7mm}
\ecol
\hspace{1mm}
\vspace{-27mm}
\bcol{0.5\textwidth}

**Fitting a JSDM** from available data in New Caledonia

\vspace{6mm}
\includegraphics[height=0.29\textheight]{figs/JSDM_NC.png}
\vspace{10mm}
\ecol
\ecols

### Obtaining biodiversity maps in New Caledonia 

#### Estimated species richness in New Caledonia 

Using the adjusted JSDM, it was possible to obtain biodiversity maps for New Caledonia following the same method as that presented for Madagascar. 

\vspace{-1mm}
\bcols
\hspace{-18mm}
\bcol{0.4\textwidth} 
\begin{center}
\includegraphics[height=0.5\textheight]{figs/species_richness_observed_vs_estimated.png}
\end{center}
\vspace{-2mm}
**Species richness** estimated and observed at **inventory sites**.
\ecol
\hspace{-20mm}
\bcol{0.6\textwidth} 
\vspace{-5mm}
\begin{center}
\includegraphics[height=0.52\textheight]{figs/estimated_species_richness_forest.png}
\end{center}
**Estimated species richness** in New Caledonia, restricted to the remaining forest cover in 2010
\ecol
\ecols

### Tree species community map in New Caledonia 

**Estimated $\beta$ diversity in New Caledonia**, restricted to the remaining forest cover in 2010

\vspace{-2mm}
\begin{center}
\includegraphics[height=0.8\textheight]{figs/RGB.png}
\end{center}

### Classification of tree communities in New Caledonia

\hspace{-3mm}
\bcols
\hspace{-6mm}
\bcol{0.55\textwidth} 
\includegraphics[height=0.75\textheight]{figs/color_group.png}
\ecol

\hspace{6mm}
\bcol{0.5\textwidth} 
\vspace{-6mm}
**Method**: 

- **t-SNE** (t-distributed Stochastic Neighbor
Embedding) on **species' occurrence probabilities**.
\vspace{1mm}
- **Hierarchical Ascending Classification** to determine the number of species communities to be considered: **6 groups** 

<!-- (determined by the first break on the dendrogram) -->
\vspace{1mm}
- **K-means classification** (K=6) performed on the coordinates of the t-SNE's three axes. 
\vspace{1mm}
- **Characterization of a group** by the 20 species with the highest and the 20 ones with the lowest average probability of occurrence on this group's pixels. 
\vspace{1mm}

\ecol
\ecols


# Fitting JSDMs in French Guiana 

## Objectives of METRADICA (Task 3)

Using JSDMs:

- Test **trait-environmment** interactions for determining tree species distribution in French Guiana.
- Assess species vulnerability to climate change (through contraction of species range).
- Interpret species vulnerability to climate change in terms of functional traits.
- Derive maps of $\alpha$ and $\beta$ diversity for French Guiana.
- Identify refuge area for biodiversity under climate change (stable tree communities).

<!-- ============================================ -->
<!-- Methods -->
<!-- ============================================ -->

## Datasets

### Datasets

\bcols
\bcol{0.5\textwidth}

Three types of data-sets:
\vspace{3mm}

- Species occurrences on sites
- Species trait database
- Environmental database

\ecol
\bcol{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/datasets2.pdf}
\ecol
\ecols

### Occurrences

```{r inventories, include=FALSE}
## Libraries
library(readr)
## Guyafor
f <- "/home/clement/Documents/projet_METRADICA/data/Inventaires/202110_DonneesModeleMetradicaTout+Diam.csv"
df <- readr::read_delim(f, delim=";")
nplots_1 <- length(unique(paste0(df$Project, df$idPlot)))
## Guyadiv
f <- "/home/clement/Documents/projet_METRADICA/data/Inventaires/20211203_GuyaDiv.csv"
df <-  readr::read_delim(f, delim=",")
nplots_2 <- length(unique(paste0(df$Forest, df$Plot)))
## Total
nplots <- nplots_1 + nplots_2
```

\bcols
\bcol{0.5\textwidth}

- Forest plot inventories coming from several networks combined together
\vspace{2mm}
- Networks: Guyafor, Gentry, Habitat, Guyadiv
\vspace{2mm}
- Presence-absence data and abundances
\vspace{2mm}
- **`r nplots` forest plots**
\vspace{2mm}
- About **1700 tree species**, most of which are rare

\ecol
\bcol{0.6\textwidth}
\includegraphics[width=\textwidth]{figs/inventories}
\ecol
\ecols

### Traits

- Large "soft" trait (WD, LSA, tree max height, etc.) databases from previous CEBA projects.
\vspace{2mm}
- Five additional mechanistic traits from Metradica project (Task 1 & 2 Marion Boisseaux and Daniela Krebber):
\vspace{2mm}
  - leaf water potential at which cells lose turgor (Ptlp), minimum leaf conductance (gmin), leaf saturated water content (LSWC), vein density (VLA), stomatal density (SD).
  \vspace{2mm}
  - 24 species, 672 trees, three sites with both hills and valleys spread on a precipitation gradient.

### Environment

**Variables:**

- Topographic data (SRTM and LiDAR, Slope, aspect, roughness, solar radiation, TWI...)
\vspace{2mm}
- Soil data
\vspace{2mm}
- Distance to human infrastructures (roads, villages...)
\vspace{2mm}
- Climatic data ([Chelsa](https://chelsa-climate.org/)) in the present and the future
\vspace{2mm}
- **PAWD**: Plant available water deficit, more relevant index to characterize hydrological constraints in each site constructed by Salman Khan (Task 1)

\vspace{2mm}

**R package to gather these data: <https://ecology.ghislainv.fr/gecevar/>**

<!-- - User friendly -->
<!-- - Allows the user to specify the desired: -->
<!--   * geographic area -->
<!--   * resolution -->
<!--   * spatial projection -->
<!-- - Efficiency:  -->
<!--   * R packages `terra` and `stars` for spatial objects -->
<!--   * Software [`GRASS GIS`](https://grass.osgeo.org/) via `rgrass` R package and [`GDAL`](https://gdal.org/) via `rgdal` R package, used for reprojection, resampling and data computation.  -->

## Study scales: biogeography and micro-environment

### Scales: biogeography and micro-environment

#### Country scale (French Guiana)

\bcols
\bcol{0.7\textwidth}

- Scale $=$ FG, resolution $=$ ~1km
\vspace{2mm}
- Explicative and predictive model
\vspace{2mm}
- Two models
  * Without traits
    - Predictive model
    - Present: distribution and co-occurrences of species
    - Future: range contraction in the future: 
    (i) species vulnerability to climate change
    (ii) change in species composition
    \vspace{2mm}
  * With traits
    - Explicative model: E $\times$ T
    - Explaining species location (biogeography)
    
\ecol
\bcol{0.3\textwidth}
\centering
\includegraphics[width=1.2\textwidth]{figs/locationParacou.jpg}
\ecol
\ecols

    
### Scales: biogeography and micro-environment

#### Local scale: microtopography $\times$ traits
\vspace{2mm}
- Scale $=$ ~10km, resolution $=$ ~10m
\vspace{2mm}
- Redrawing of Paracou's plots into $10 \times 10$ m sites 
\vspace{2mm}
- Explicative model: E $\times$ T
\vspace{2mm}
- Environmental variable considered:
\vspace{2mm}
  * MNT at 10m: hills (_"terra firme"_) and valleys
  * Topographic water deficit (TWI)
  * Plant available water deficit (PAWD, Task 1) 

\vspace{2mm}
- Species traits considered: 
\vspace{2mm}
  * Water use traits (Task 1 & 2) available for only few species 
  * Soft traits available for more species for fitting another JSDM considering mores species. 

## Ecological interest 

### Ecological interest 

- Anticipating climate change effects on tropical forest in French Guiana
\vspace{3mm}
  - Massive tree mortality events and forest conversion ?
  \vspace{2mm}
  - Change in species composition?
  
\vspace{3mm}

- Identification of refuge areas for conservation $\Rightarrow$ systematic conservation planning.

# Planned publications

- **Rcpp* package `jSDM` functionalities**: Main features and functionality of the R package `jSDM` and comparison with other approaches for the estimation of joint species distribution models. (Methods in Ecology and Evolution)

- **Fitting JSDMs with species traits in French Guiana**: Identification of tree species vulnerable to climate change and of biodiversity refuge areas in French Guiana through the use of a joint species distribution model integrating trait-environment interactions.

- **R package `gecevar` to download environmental and climatic data** : facilitate and automate the extraction and formatting of environmental and climatic data required for the adjustment of JSDMs.

- **Improve convergence of JSDMs including latent variables** : Diagnosis of latent axes and estimation of their number in generalized linear latent variable models for joint species distribution modeling (first author: Frederic Gosselin).

### References {.allowframebreaks}

