---
classoption: table, dvipsname, compress
bibliography:  ref_GAMBAS.bib
biblio-style: "apalike"
output: 
  beamer_presentation:
    # keep_tex: true
    # toc: true
    slide_level: 3
    includes:
      in_header: header.tex
      before_body: doc_prefix.tex
      after_body: doc_suffix.tex
colorlinks: yes
fontsize: 10pt
link-citations: yes
linkcolor: Black
urlcolor: Maroon
---


```{r options, include=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(echo=TRUE, cache=TRUE,
               #results="hide", 
               warning=FALSE,
               message=FALSE, highlight=TRUE,
               fig.show="hide", size="small",
               fig.align="center",
               tidy=FALSE)
options(knitr.kable.NA="-")
```


# Introduction

## JSDMs

### Joint Species Distribution Models (JSDMs)

**Species Distribution Model (SDM), for one single species.**

- $\red{y_i} \sim \mathcal{B}ernoulli( \blue{\theta_i} )$
- $i$: site
- $g(\theta_i) = \red{X_i} \blue{\beta}$
- $\red{X}$: environmental variables
- $\blue{\beta}$: species effects

\begin{center}
\includegraphics[width=0.6\textwidth]{figs/SDM.jpg}
\end{center}

**JSDM $=$ SDM for community of species.**

### Joint Species Distribution Models (JSDMs)
**Model definition**

- $\red{y_{ij}}  \sim \mathcal{B}inomial(\red{n_i} , \blue{\theta_{ij}})$, for presence-absence data,
\vspace{0.5mm}
- $\red{y_{ij}} \sim\mathcal{P}oisson(\blue{\theta_{ij}})$, for abundance data, such that: 
\vspace{-0.7mm}
   $$g(\blue{\theta_{ij}})= \blue{\alpha_i} + \red{X_i}\blue{\beta_j} + \blue{W_i\lambda_j}$$
- $i$: site, $j$: species
- $g$: a link function (probit, logit or log).
- $\red{n_i}$: number of visits to site $i$.
- $\blue{\theta_{ij}}$: occurrence probability or mean abundance of species $j$ at site $i$.
- $\blue{\alpha_i}$: site effect (mean site suitability)
- $\red{X_i}$: environmental variables including the intercept.  
- $\blue{\beta_j}$: species effect  for species $j$,
- $\blue{W_i}$: latent variables (or "unmeasured predictors") for site $i$,
- $\blue{\lambda_j}$: factor loadings for species $j$
- $\blue{\Sigma_{jg}}=\blue{\lambda_j^t\lambda_g}$: coefficient for species $j$ and $g$ of the variance-covariance matrix  (species co-occurrences) used to compute the **full species residual correlation matrix**:
  $\blue{R_{jg}}:=\frac{\blue{\Sigma_{jg}}}{\sqrt{\blue{\Sigma _{jj}\Sigma _{gg}}}}.$

### Joint Species Distribution Models

JSDMs provide a convenient statistical framework to test **trait-environment** interactions.
\vspace{2mm}
\bcols
\bcol{0.7\textwidth}

$\beta_j$ can be expressed as a function of functional traits

- if $n$ species traits are considered:
    $\blue{\beta_j} \sim \mathcal{N}_{p+1}(\blue{\mu_{\beta_j}},V_{\beta})$,
    * with $\blue{\mu_{\beta_{jk}}}  = \sum_{r=0}^{n}  \blue{T_{jr}}.\blue{\gamma_{rk}}$, for $k \in \llbracket 0,p \rrbracket$, considering $p$ covariates. 
    * where $\blue{\gamma}=\bordermatrix{ & Int & X_1 & \ldots & X_p \cr
      Int & \blue{\gamma_{0,0}} & \blue{\gamma_{0,1}} & \ldots & \blue{\gamma_{0,p+1}} \cr
      T_1 & \blue{\gamma_{1,0}} & \blue{\gamma_{1,1}} & \ldots & \blue{\gamma_{1,p+1}} \cr
      \vdots & \dots & \ldots & \ldots & \ldots \cr
      T_n & \blue{\gamma_{n,0}} & \blue{\gamma_{n,1}} & \ldots & \blue{\gamma_{n+1,p+1}}}$
      
\vspace{2mm}   

- else:  $\blue{\beta_j} \sim \mathcal{N}_{p+1}(\mu_{\beta},V_{\beta})$, for $j \in  \llbracket 1,J \rrbracket$.

\ecol 

\bcol{0.3\textwidth}

\centering 

\includegraphics[width=0.9\textwidth]{figs/four-corner-model}
\ecol

\ecols

JSDMs can help narrow the gap between **correlative** and **mechanistic** species distribution models.

## METRADICA's objectives

### Objectives of METRADICA (Task 3)

Using JSDMs:

- Test **trait-environmment** interactions for determining tree species distribution in French Guiana.
- Assess species vulnerability to climate change (through contraction of species range).
- Interpret species vulnerability to climate change in terms of functional traits.
- Derive maps of $\alpha$ and $\beta$ diversity for French Guiana.
- Identify refuge area for biodiversity under climate change (stable tree communities).

<!-- ============================================ -->
<!-- Methods -->
<!-- ============================================ -->

# Material and methods

## Datasets

### Datasets

\bcols
\bcol{0.5\textwidth}

Three types of data-sets:
\vspace{3mm}

- Species occurrences on sites
- Species trait database
- Environmental database

\ecol
\bcol{0.5\textwidth}
\centering
\includegraphics[width=\textwidth]{figs/datasets2.pdf}
\ecol
\ecols

### Occurrences

```{r inventories, include=FALSE}
## Libraries
library(readr)
## Guyafor
f <- "/home/clement/Documents/projet_METRADICA/data/Inventaires/202110_DonneesModeleMetradicaTout+Diam.csv"
df <- read_delim(f, delim=";")
nplots_1 <- length(unique(paste0(df$Project, df$idPlot)))
## Guyadiv
f <- "/home/clement/Documents/projet_METRADICA/data/Inventaires/20211203_GuyaDiv.csv"
df <-  read_delim(f, delim=",")
nplots_2 <- length(unique(paste0(df$Forest, df$Plot)))
## Total
nplots <- nplots_1 + nplots_2
```

\bcols
\bcol{0.5\textwidth}

- Forest plot inventories coming from several networks combined together
- Networks: Guyafor, Gentry, Habitat, Guyadiv
- Presence-absence data and abundances
- `r nplots` forest plots
- About 1700 tree species, most of which are rare

\ecol
\bcol{0.5\textwidth}
\includegraphics[width=\textwidth]{figs/inventories}
\ecol
\ecols

### Traits

- Large "soft" trait (WD, LSA, tree max height, etc.) databases from previous CEBA projects.
- Five additional mechanistic traits from Metradica project:
  - leaf water potential at which cells lose turgor (Ptlp), minimum leaf conductance (gmin), leaf saturated water content (LSWC), vein density (VLA), stomatal density (SD).
  - 24 species, 672 trees, three sites with both hills and valleys spread on a precipitation gradient.

### Environment

**Variables:**

- Topographic data (SRTM and LiDAR, Slope, aspect, roughness, solar radiation, TWI...)
- Soil data
- Distance to human infrastructures (roads, villages...)
- Climatic data ([Chelsa](https://chelsa-climate.org/)) in the present and the future

**R package to gather these data: <https://guyaclim.cirad.fr>:**

- User friendly
- Allows the user to specify the desired:
  * geographic area
  * resolution
  * spatial projection
- Efficiency: 
  * R packages `terra` and `stars` for spatial objects
  * Software [`GRASS GIS`](https://grass.osgeo.org/) via `rgrass` R package and [`GDAL`](https://gdal.org/) via `rgdal` R package, used for reprojection, resampling and data computation. 

## Study scales

### Scales: biogeography and micro-environment

\bcols
\bcol{0.7\textwidth}

**Local scale: microtopography $\times$ traits**

- Scale $=$ ~10km, resolution $=$ ~5m
- Explicative model: E $\times$ T
- Using MNT at 5m: hills (_"terra firme"_) and valleys

**Country scale (French Guiana)**

- Scale $=$ FG, resolution $=$ ~1km
- Explicative and predictive model
- Two models
  * Without traits
    - Predictive model
    - Present: distribution and co-occurrences of species
    - Future: range contraction in the future: (i) species vulnerability to climate change, (ii) change in species composition
  * With traits
    - Explicative model: E $\times$ T
    - Explaining species location (biogeography)

\ecol
\bcol{0.3\textwidth}
\includegraphics[width=\textwidth]{figs/locationParacou.jpg}
\ecol
\ecols

## `jSDM` R package

### `jSDM` R package

\includegraphics[width=0.97\textwidth]{figs/jSDM-Website.png}

- <https://ecology.ghislainv.fr/jSDM>
- Made with Rcpp* packages
<!-- ============================================ -->
<!-- Perspectives -->
<!-- ============================================ -->

## Benchmarking our method against other R packages 

### Benchmarking our method against other R packages 

#### R packages compared

- `Hmsc` 3.0-11:  R package for Hierarchical Modelling of Species Communities (based on R code).
- `boral` 2.0:  R package for Bayesian Ordination and Regression Analysis (based on `JAGS` code <http://mcmc-jags.sourceforge.net/>).
- `jSDM` 0.2.1 : R package, we have developed (based on Rcpp and C++ compiled code). 

\pause
#### Data-sets used 

- 4 real presence-absence data-sets, used in a recent JSDM benchmark study by @Wilkinson2019 and a simulated one (binomial probit regressions)

- 2 real abundance data-sets [@Choler2005; @Borcard1994] (Poisson log-linear regressions)

\pause
#### Model definition

We fitted JSDMs of the form: $g(\blue{\theta_{ij}}) = \red{X_i}\blue{\beta_j} + \blue{W_i\lambda_j}$. 

### Computation times and accuracy of results 

\begin{center}
\includegraphics[width=0.8\textwidth]{figs/comparison_poster.pdf}
\end{center}

- RMSE of the occurrence probabilities ($\theta_{ij}$) for the simulated data-set. 
\vspace{2mm}
- Deviance computed as follows : $D=-2\sum_{i}\sum_j\log(\mathbb {P}(y_{ij}\ |\ \hat{\theta}_{ij}))$
\vspace{2mm}
- TSS: True Skills Statistic, TSS$:=$sensitivity$+$specificity$-1$.

### Comparison of computation times and accuracy of results 

\vspace{4mm}
- `jSDM` is **2** to **7** times faster than `Hmsc` and **15** to **439** times faster than `boral` for presence-absence data-sets.
\vspace{4mm}
- `jSDM` is **1.5** to **2.1** times faster than `Hmsc` and **1.7** to **2.9** times faster than `boral` for abundance data-sets. 
\vspace{4mm}
- `jSDM` always provides either equivalent or more accurate estimates than `Hmsc`, given the RMSE, TSS and Deviance values. 

# Perspectives

## Model comparison

### Model comparison with forest dynamics models

**TROLL model**

- Tropical forest dynamics model
- Growth, mortality, recruitment through carbon allocation
- Species parameters are derived from traits
- Calibrated on some forests of French Guiana

**Model comparison**

- Species excluded from the community with TROLL under climate change.
- Do the same species experience a severe range contraction with JSDMs?

## Applications

### Applications

- Anticipating climate change effects on tropical forest in French Guiana
\vspace{3mm}
  - Massive tree mortality events and forest conversion to savannas?
  \vspace{2mm}
  - Change in species composition?
  
\vspace{3mm}

- Identification of refuge areas for conservation $\Rightarrow$ systematic conservation planning.

### References {.allowframebreaks}


<!-- ============================================ -->
<!-- EOF
<!-- ============================================ -->
